```{r, echo = F, include = F}

# Load all necessary R Packages first

packages <- c("tidyverse", "latex2exp", "gridExtra", "rvest", "kableExtra")
lapply(packages, require, character.only = T)
```
unbiased: $E(T) = \theta$
consistency: $T_n \rightarrow ^p \theta$ as $n \rightarrow \infty$ means estimator on larger and larger sample sizes will reflect the true param. perhaps the estimator is biased. but it could still be consistent as sample size inc. 

to show biasedness in r, calculate point estimate of the distribution with estimator and increasing sample sizes. 5, 10, 15, 100. then graph all point estimate, which will show $E(T)$. compare with true $\theta$. if they are the same val, then unbiased. 

```{r}
# let's estimate 1 realization of some pois distribution with param lambda, and we estimate it with lambda hat to be a sample mean of size 10. 
simulated_x <- rpois(n = 10, lambda = 1)
point_estimate_lambda_hat <- mean(simulated_x)
real_lambda <- exp(-1)

# many realizations, check E(T) as well
m <- 5000 # number of copies to create
t1 <- numeric(m) # an empty numeric vector of size m
for (i in (1:m)) {
  simulated_x <- rpois(10, lambda = 1)  # calculate sample mean w size ten and randomly generated samples with lambda = 1 5000 times
  t1[i] <- exp(-mean(simulated_x)) #store all point estimates in vec
}
mean(t1) # gets expected val of estimator 
# expected val of T is pretty close to real_lambda but there is bias, expected val is larger.
```
check consistency: increase n from 10 to 1000 in increments of 5, calculate E(T) each time
```{r}
m <- 5000
ns <- seq(10, 1000, 10)
t1_means <- numeric(length(ns))
t1 <- c(length(m))
for (j in seq(length(ns))) { # calculate w sample sizes as computed by ns sequence
  t1 <- numeric(m)
  for (i in seq(m)) { # calculate each sample size 5000 times 
    simulated_x <- rpois(n = ns[j], lambda = 1) 
    t1[i] <- exp(-mean(simulated_x)) 
  }
  t1_means[j] <- mean(t1) # compute each point estimate's mean
}
```
you will see that E(T) converges to the true parameter as ns values get larger. consistency. 